Blum axioms and nondeterministic computation of functions

### 2016-04-17 paper.tex
paper.tex: Rewrite the presentation od Goldreich's definition.

The extra symbol $\bot$ sidesteps the possibility of a branch looping forever.
Now,
every branch is required to halt for the value of the function be defined.
Thus,
we can convert a nondeterministic Turing machine to a deterministic machine
that compute the same function,
albeit with an exponential slowdown
--- the deterministic machine just need to simulate all branches until completion,
to actually be sure every branch halts.

This allows us to comply with the requirements to being an acceptable Gödel numbering:
\begin{itemize}
    \item Any deterministic machine can be understood as a nondeterministic machine
        with a single branch,
        so every computable function appears in the corresponding numbering.
        Thus, the numbering is surjective --- it enumerates every computable function.
    \item Any nondeterministic machine can be simulated by some deterministic machine,
        so the enumeration codomain is the set of computable functions.
    \item This conversion can be done mechanically,
        so we can use the same universal machine and $\sigma$ function
        of the standard Gödel numbering $\pi$
        to comply with the universal machine theorem and the $S_{mn}$ theorem.
\end{itemize}

At least partially,
we can compute the characteristic function of the satisfiability problem
in linear time using this model.
The machine $M$ that will do the calculation will guess an assignment
and write $1$ if it is satisfying and write $\bot$ if it isn't.
This computes in linear time the function $f$, defined by
\begin{equation*}
    f(\varphi) = \begin{cases}
        1, & \text{if $\varphi$ is satisfiable.} \\
        \text{undefined}, & \text{if $\varphi$ is unsatisfiable.} \\
    \end{cases}
\end{equation*}
However,
this function falls short of being exactly $\SAT$'s characteristic function:
$f(\varphi)$ ought to be zero for unsatisfiable $\varphi$.
Although we cannot prove a strong assertion as
``if this works then $\P = \NP$'',
we can come close.

@@

($\NP = \coNP$ implies that $\NP = \PH$; that is,
the polynomial hierarchy collapses to the first level.)

@@

    To solve $\overline \SAT$,
    the complement of $\SAT$,
    using $\NP$ resources,
    create a nondeterministic machine $M'$ that will simulate $M$ on its input.
    On each branch, if $M$ returned either $\bot$ or $1$, reject;
    if $M$ returns $0$, accept.

    If the current instance is satisfiable,
    all branches will return either $\bot$ or $1$,
    so we are right in rejecting it;
    otherwise, if the current instance is unsatisfiable,
    at least some branch will return $0$,
    so we will correctly reject the input.

    Since the complement of $\SAT$ is $\coNP$-complete,
    any language in $\coNP$ can thus be solved with a nondeterministic machine;
    hence, $\coNP \subseteq \NP$.

    Similarly,
    every language $L$ of $\NP$ has a many-one reduction to $\SAT$;
    thus, the complement $\overline L$ has a many-one reduction
    to $\overline \SAT$.
    Hence, composing the reduction with the above algorithm
    that solves $\overline \SAT$ using a nondeterministic machine,
    we conclude that $\overline L \in \NP$.
    Thus, by definition, $L \in \coNP$.
    So, $\NP \subseteq \coNP$.

    This shows $\NP = \coNP$.
\end{proof}

Therefore, even though Goldreich's definition
yields an acceptable Gödel numbering,
it also have trouble in transposing the exponential speedup
we have using nondeterminism.

>>>>

The extra symbol $\bot$ sidesteps the problems of a branch looping forever.
Now,
every branch is required to halt for the value to be defined.
Thus,
we can convert a nondeterministic Turing machine to a deterministic machine
that compute the same function,
albeit with an exponential slowdown
--- the deterministic machine just need to simulate all branches until completion,
to actually be sure every branch halts;
if some branch do not halt,
then the simulating machine will not halt either,
but the function is not defined in this case,
so this behavior is correct.

Also, a deterministic machine functions exactly the same way under Goldreich's definition,
so we can convert mechanically
between the numbering corresponding to this definition
and a standard numbering for deterministic machines.
Therefore,
to show the universal machine theorem and the $S_{mn}$ theorem,
we can first convert the machine in question to a deterministic machine
and use their theorems.
Thus, this definition comply with the requirements to being an acceptable Gödel numbering,

To establish a complexity measure,
we can count the number of steps of a branch with most steps,
and leave it undefined if some branch loops forever.
--- this is the same method
we use to evaluate the time complexity of a nondeterministic decider.
Leaving undefined if a branch loops forever guarantees the first Blum axiom is satisfied.
And, to decide the predicate of the second axiom,
we can simply run all the branches of the machine for the specified number of steps.
Every branch has to halt within that number of steps,
and some branch must take exactly that number,
for the predicate to be true.

So, Goldreich's definition defines an acceptable Gödel numbering
and we can form a complexity measure that satisfies Blum axioms.
But, again,
we have trouble with the ``exponential speedup'' requirement.

At least partially,
we can compute the characteristic function of the satisfiability problem
in linear time using this model.
The machine $M$ that will do the calculation will guess an assignment
and write $1$ if it is satisfying and write $\bot$ if it isn't.
This computes in linear time the function $f$, defined by
\begin{equation*}
    f(\varphi) = \begin{cases}
        1, & \text{if $\varphi$ is satisfiable.} \\
        \text{undefined}, & \text{if $\varphi$ is unsatisfiable.} \\
    \end{cases}
\end{equation*}
However,
this function falls short of being exactly $\SAT$'s characteristic function:
$f(\varphi)$ ought to be zero for unsatisfiable formulas $\varphi$.
Although we cannot prove a strong assertion as
``if this works then $\P = \NP$'',
we can come close.

@@

($\NP = \coNP$ implies that $\NP = \PH$; that is,
the polynomial hierarchy collapses to the first level \cite[p.~280]{Kozen2006}.)

@@

    The characteristic function of $\SAT$ is a total function,
    and its only outputs are $0$ and $1$.
    Therefore, in every computation of $M$,
    there is at least one branch that writes something different that $\bot$ on the tape,
    and whatever it writes is the correct answer.
    So, if we convert this to a nondeterministic decider and invert the output
    (branches that write $0$ will accept the input and vice-versa;
    branches that write $\bot$ always reject),
    any satisfiable formula will be rejected,
    because no branch of $M$ ever writes $0$ on the tape for these formulas,
    and any unsatisfiable formula will be accepted,
    because at least one branch of $M$ writes $0$ for these formulas.
    we can decide $\overline \SAT$, the complement of $\SAT$.

    Since $\overline \SAT$ is $\coNP$-complete,
    the existence of such a machine $M$
    would show that $\coNP \subseteq \NP$,
    and this implies that $\coNP = \NP$.\footnote{
        To see why $\coNP \subseteq \NP$ implies $\coNP = \NP$,
        pick a language $L \in \NP$.
        Its complement $\overline L$ is in $\coNP$, by the definition of $\coNP$.
        But by hypothesis, $\overline L$ is in $\NP$,
        so, by the definition of $\coNP$,
        its complement, $\overline{\overline L} = L$ is in $\coNP$,
        thus showing $\coNP = \NP$.
    }
\end{proof}

Therefore, even though Goldreich's definition
yields an acceptable Gödel numbering and a Blum complexity measure,
it also have trouble in transposing the exponential speedup
we have using nondeterminism.
